---
title: "Get started with allometree"
author: 
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Get started with allometree}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE, warning = FALSE
)

devtools::load_all()
#library(allometree) # if installed
```

The `allometree` package enables you to develop and use allometric equations relating to the size and structure of urban trees. For example, these equations have been used to predict trunk diameter from tree age, as well as to predict tree height, crown height, crown diameter and leaf area from trunk diameter. They are foundational to other models that estimate the benefits and hazards associated with trees as they mature and grow in size. More information can be found in Song *et al.* (in prep).  

This document demonstrates the workflow to develop these allometric models, taking the relationship between two size parameters as an example: Predicting tree *height* from trunk *diameter*. 

Two linear modelling approaches will be introduced:  

  1. Regression models developed separately for each species, i.e., *single-species models*
  2. Mixed-effects model that includes all species as random effects, i.e., *mixed-effects model*


## Example data: `urbantrees`

We will be using `data(urbantrees)`. It contains five species planted along streets in Singapore, each spanning a wide range of heights and diameter sizes (in metres).

```{r load data, fig.width=5.5, fig.align='center'}
data(urbantrees, package = "allometree")
urbantrees

library(ggplot2)
ggplot(urbantrees, aes(diameter, height)) +
  facet_wrap(~ species, scales = "free") +
  geom_point()
```


## Allometric equations: `eqns_info`

Allometric relationships for urban trees can vary drastically from those for forest trees, and are also influenced by human factors such as pruning and fertilisation. Empirical models are developed using six allometric equations used for urban trees (McPherson *et al.*, 2016). 

More information can be found in `?eqns_info`, as well as in `data(eqns_info)`. The column `modelcode` shows the unique model code for each equation:

```{r eqns_info}
data(eqns_info, package = "allometree")
head(eqns_info)
```

---

## 1. Single-species models

### 1.1. Model selection {#section1.1}

#### For one species

For a particular species of interest, data will be fit to all allometric equations. The best-fit model is selected based on the lowest bias-corrected Aikaike’s information criterion (AICc) value. Here's an example to select the best-fit model for the species *Albizia saman*, using the function `sp_modelselect()`.

```{r sp_modelselect}
Alb_sam <- urbantrees[urbantrees$species == "Albizia saman", ]  # subset data for one species

results <- sp_modelselect(Alb_sam, 
                          response = "height", predictor = "diameter")  # specify colnames of variables
```

The `results` variable is a list of 3 elements:

1. Table showing models ranked by AICc value, named `all_models_rank`.
2. Best-fit model object, named `best_model`.
3. Table showing information on the best-fit model, named `best_model_info`. More information on the output columns can be found at `?sp_modelselect`.

```{r}
head(results$all_models_rank)
```

```{r}
results$best_model
```

```{r}
results$best_model_info
```


#### For multiple species

We can also run `sp_modelselect()` across multiple species in the full dataset `urbantrees`. For this, we use the wrapper function `sp_modelselect_multi()`.

```{r}
results_all <- sp_modelselect_multi(urbantrees, species = "species", # specify colname of species
                                    response = "height", predictor = "diameter")
```

The `results_all` variable is a list of 3 elements:

1. List of tables showing each species' candidate models ranked by AICc value, named `sp_models_rank`.
2. List of each species' best-fit model object, named `sp_models`.
3. Table showing each species' best-fit model information, named `sp_models_info`. 

&nbsp;

An overview of the best-fit models across the `r nrow(results_all$sp_models_info)` species:
```{r}
results_all$sp_models_info
```


### 1.2. Outlier removal

Let's say we want to check for outlier/influential data points, remove them, and re-fit the filtered dataset to the pre-selected models as defined in `results_all$sp_models_info`. First, lets visualise these outliers, say, for the species *Xanthostemon chrysanthus*.

```{r fig.align='center', fig.width=5.5, fig.height=5.5}
par(mfrow=c(2,2)); plot(results_all$sp_models$`Xanthostemon chrysanthus`)
```
&nbsp;

There are numerous ways to [deal with outliers](http://r-statistics.co/Outlier-Treatment-With-R.html). One way is to remove them based on Cook's Distance (bottom-right plot). In general use, outliers may be defined as points with a Cook's distance more than four times the mean Cook's distance (red line in plot below). Note that this threshold is not fixed, and may be adjusted according to expert knowledge.

```{r, fig.align='center', fig.width=4, fig.height=4}
cooks_dist <- cooks.distance(results_all$sp_models$`Xanthostemon chrysanthus`)
plot(cooks_dist, pch = ".", cex = 2, main = "Outliers by Cook's Distance")
abline(h = 4*mean(cooks_dist, na.rm = T), col = "red")  # threshold line
text(x = 1:length(cooks_dist) + 1, y = cooks_dist, 
     labels = ifelse(cooks_dist > 4 * mean(cooks_dist, na.rm=T), names(cooks_dist), ""), col = "red")  # outlier labels
```

We can create a function `uncooker()` that removes outliers for each species in the dataset `urbantrees`, based on their respective best-fit models in `results_all$sp_models`:

```{r}
uncooker <- function(data, modellist){
  
  result <- data[0,]
  
  for(i in 1:length(modellist)){ # loop over all species 
    
    cooks_dist <- cooks.distance(modellist[[i]]) 
    outliers <- as.numeric(which(cooks_dist > 4*mean(cooks_dist, na.rm=T)))  # row numbers
    subset <- data[(data$species == names(modellist[i])),][-outliers,]
    result <- rbind.data.frame(result, subset)
  }
  return(result)
}

# run the function uncooker()
urbantrees_clean <- uncooker(urbantrees, results_all$sp_models)

nrow(urbantrees_clean)
```

`r nrow(urbantrees) - nrow(urbantrees_clean)` data points were removed to obtain the filtered dataset `urbantrees_clean` (n = `r nrow(urbantrees_clean)`).

```{r include = FALSE}
rm(cooks_dist, uncooker, outliers)
```


### 1.3. Re-fit filtered data

#### For one species

We can now fit `urbantrees_clean` to the best-fit equations we defined in [Section 1.1](#section1.1). Here's an example for the species *Albizia saman* again, using the function `sp_modelfit()`. 

This function has an additional argument `modelcode`. In the case of *Albizia saman*, it is `lin_w1` (see `results_all$sp_models_info`). Note that you can also pick any one of the options found in `eqns_info$modelcode`.

We'll overwrite our previously-defined variables, this time with the filtered dataset:

```{r fit data to pre-selected model, eval = FALSE}
Alb_sam <- urbantrees_clean[urbantrees_clean$species == "Albizia saman", ]

results <- sp_modelfit(Alb_sam, modelcode = "lin_w1", # specify modelcode
                       response = "height", predictor = "diameter") 
```

The new `results` variable is a list of 2 elements:

1. Resulting model object, named `fitted_model`.
2. Table showing information on the resulting model, named `fitted_model_info`.

#### For multiple species

Similar to the the wrapper function `sp_modelselect_multi()`, we can also run  `sp_modelfit_multi()` to fit pre-selected models across multiple species. In this function, we need to input a reference table `ref_table` (i.e. `results_all$sp_models_info`) that provides information on the `species` and their corresponding `modelcode`.

```{r}
results_all <- sp_modelfit_multi(urbantrees_clean, 
                                 ref_table = results_all$sp_models_info,
                                 species = "species", # colname in both data & ref_table
                                 modelcode = "modelcode", # colname in ref_table
                                 response = "height", predictor = "diameter")
```

The new `results_all` variable is a list of 2 elements:

1. List of each species' resulting model object, named `sp_models`.
2. Table showing each species' resulting model information, named `sp_models_info`.

&nbsp;

To make our subsequent code less verbose, let's assign the elements `sp_models` and `sp_models_info` to variables with the same name:

```{r}
sp_models <- results_all$sp_models
sp_models_info <- results_all$sp_models_info
```

```{r include = FALSE}
rm(results_all, results, Alb_sam)
```

### 1.4. Make predictions

In the works...

---

## 2. Mixed-effects model

### 2.1. Model selection

In the works...

---

## References

McPherson E. G., van Doorn N. S. & Peper P. J. (2016) Urban Tree Database and Allometric Equations. *General Technical Report PSW-GTR-253, USDA Forest Service*, 86.

Song, X. P., Lai, H. R., Wijedasa, L. S., Yee, A. T. K., Tan, P. Y., Richards, D. R., Streamlining management practices based on the size allometry of tropical street trees (in prep).

