---
title: "Single-species linear models"
author: 
date: "`r Sys.Date()`"
opengraph:
  image: 
    src: "man/figures/logo.png"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Single-species linear models}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE, warning = FALSE
)
```

```{r include = FALSE, eval = T}
library(allometree)
#devtools::load_all() # to knit manually
```

---

## 1. Model selection {#section1}

In this document, we are using relationship between two size parameters as an example: Predicting tree *height* from trunk *diameter*. See the vignette [Get started with allometree](allometree.html) for a full description of the dataset and allometric equations.

### For one species

For a particular species of interest, data will be fit to all allometric equations. The best-fit model is selected based on the lowest bias-corrected Aikaike’s information criterion (AICc) value. Here's an example to select the best-fit model for the species *Albizia saman*, using the function `sp_modelselect()`.


```{r sp_modelselect}
data(urbantrees, package = "allometree")

Alb_sam <- urbantrees[urbantrees$species == "Albizia saman", ]  # subset data for one species

results <- sp_modelselect(Alb_sam, 
                          response = "height", predictor = "diameter")  # specify colnames of variables
```

`results` is a list of 3 elements. More information on the output can be found at `?sp_modelselect`.

1. Table showing models ranked by AICc value, named `all_models_rank`.
2. Best-fit model object, named `best_model`.
3. Table showing information on the best-fit model, named `best_model_info`. 

```{r}
head(results$all_models_rank)
```

```{r}
results$best_model
```

```{r}
results$best_model_info
```


### For multiple species

We can also run `sp_modelselect()` across multiple species in the full dataset `urbantrees`. For this, we use the wrapper function `sp_modelselect_multi()`.

```{r}
results_all <- sp_modelselect_multi(urbantrees, species = "species", # specify colname of species
                                    response = "height", predictor = "diameter")
```

`results_all` is a list of 3 elements:

1. List of tables showing each species' candidate models ranked by AICc value, named `sp_models_rank`.
2. List of each species' best-fit model object, named `sp_models`.
3. Table showing each species' best-fit model information, named `sp_models_info`. 

&nbsp;

An overview of the best-fit models across the `r nrow(results_all$sp_models_info)` species:
```{r}
results_all$sp_models_info
```


## 2. Outlier removal

Let's say we want to check for outlier/influential data points, remove them, and re-fit the filtered dataset to the pre-selected models as defined in `results_all$sp_models_info`. First, lets visualise these outliers, say, for the species *Xanthostemon chrysanthus*.

```{r fig.align='center', fig.width=5.5, fig.height=5.5}
par(mfrow=c(2,2)); plot(results_all$sp_models$`Xanthostemon chrysanthus`)
```
&nbsp;

There are numerous ways to [deal with outliers](http://r-statistics.co/Outlier-Treatment-With-R.html). One way is to remove them based on Cook's Distance (bottom-right plot). In general use, outliers may be defined as points with a Cook's distance more than four times the mean Cook's distance (red line in plot below). Note that this threshold is not fixed, and may be adjusted according to biological knowledge of each tree species.

```{r, fig.align='center', fig.width=4, fig.height=4}
cooks_dist <- cooks.distance(results_all$sp_models$`Xanthostemon chrysanthus`)
plot(cooks_dist, pch = ".", cex = 2, main = "Outliers by Cook's Distance")
abline(h = 4 * mean(cooks_dist, na.rm = T), col = "red")  # threshold line 
text(x = 1:length(cooks_dist) + 1, y = cooks_dist, 
     labels = ifelse(cooks_dist > 4 * mean(cooks_dist, na.rm=T), names(cooks_dist), ""), col = "red")  # outlier labels
```

&nbsp;

Lets remove outliers across all species in `urbantrees`. We can create a function `uncooker()` that does so for each species using a `for()` loop, based on their respective best-fit model in `results_all$sp_models`:

```{r}
uncooker <- function(data, modellist){
  
  result <- data[0,]
  
  for(i in 1:length(modellist)){ # loop over all species 
    
    cooks_dist <- cooks.distance(modellist[[i]]) 
    outliers <- as.numeric(which(cooks_dist > 4 * mean(cooks_dist, na.rm=T)))  # threshold
    subset <- data[(data$species == names(modellist[i])),][-outliers,]
    result <- rbind.data.frame(result, subset)
  }
  return(result)
}

# run the function uncooker()
urbantrees_clean <- uncooker(urbantrees, results_all$sp_models)

nrow(urbantrees_clean)
```

`r nrow(urbantrees) - nrow(urbantrees_clean)` data points were removed to obtain the filtered dataset `urbantrees_clean` (n = `r nrow(urbantrees_clean)`).

```{r include = FALSE}
rm(cooks_dist, uncooker)
```


## 3. Re-fit filtered data

### For one species

We can now fit `urbantrees_clean` to the best-fit equations we defined in [Section 1](#section1). Here's an example for the species *Albizia saman* again, using the function `sp_modelfit()`. This function has an additional argument `modelcode`. In the case of *Albizia saman*, it is `lin_w1` (see `results_all$sp_models_info`). Note that you can also pick any one of the options found in `eqns_info$modelcode`.

&nbsp;

We'll overwrite our previously-defined variables `Alb_sam` and `results`, this time with the filtered dataset:

```{r fit data to pre-selected model, eval = FALSE}
Alb_sam <- urbantrees_clean[urbantrees_clean$species == "Albizia saman", ]

results <- sp_modelfit(Alb_sam, modelcode = "lin_w1", # specify modelcode
                       response = "height", predictor = "diameter") 
```

`results` is now a list of 2 elements:

1. Resulting model object, named `fitted_model`.
2. Table showing information on the resulting model, named `fitted_model_info`.

### For multiple species

Similar to the the wrapper function `sp_modelselect_multi()`, we can also run  `sp_modelfit_multi()` to fit pre-selected models across multiple species. In this function, we need to input a reference table `ref_table` (i.e. `results_all$sp_models_info`) that provides information on the `species` and their corresponding `modelcode`.

```{r}
results_all <- sp_modelfit_multi(urbantrees_clean, 
                                 ref_table = results_all$sp_models_info,
                                 species = "species", # colname in both data & ref_table
                                 modelcode = "modelcode", # colname in ref_table
                                 response = "height", predictor = "diameter")
```

`results_all` is now a list of 2 elements:

1. List of each species' resulting model object, named `sp_models`.
2. Table showing each species' resulting model information, named `sp_models_info`.

&nbsp;

To make our subsequent code less verbose, let's assign the elements `sp_models` and `sp_models_info` to variables with the same name:

```{r}
sp_models <- results_all$sp_models
sp_models_info <- results_all$sp_models_info
```

```{r include = FALSE}
rm(results_all, results, Alb_sam)
```

## 4. Make predictions

We can simulate some data for a species of interest, and use it's model to make predictions. We'll use the species *Albizia saman* again as an example. Let's first simulate the data, based on the range of values for the predictor variable (i.e. diameter size) that was used to fit the model:

```{r}
predict_range <- sp_models_info[sp_models_info$species == "Albizia saman",]

# simulate 100 data points across range
predict_range_full <- as.data.frame(apply(predict_range, 1, function(x) seq(x["predictor_min"], x["predictor_max"], length.out = 100)))
colnames(predict_range_full) <- predict_range$species

predict_range_full <- tidyr::pivot_longer(predict_range_full, 
                                          cols = colnames(predict_range_full), 
                                          names_to = "species", values_to = "predictor")
```

&nbsp;

`sp_predict()` can be used to make predictions on the simulated data `predict_range_full`. This appends columns for predicted height (`fit`), as well as the lower (`lwr`) and upper (`upr`) bounds of the prediction interval:

```{r}
predictions <- sp_predict(predict_range_full, 
                          models = sp_models, 
                          ref_table = sp_models_info, 
                          predictor = "predictor")
head(predictions)
```

&nbsp;

Note that you can simulate data automatically using the `sp_simulate()` function, which is also a wrapper to `sp_predict()`. This can be done for one or multiple species, using the argument `select_sp`. The simulated data can be extrapolated beyond the range used to fit the model, using the argument `extrapolate`. Here's an example using the models that we previously derived from the filtered dataset `urbantrees_clean`. Let's say we're interested in *Albizia saman*, *Terminalia mantaly*, and *Xanthostemon chrysanthus*, and would like to predict their height across a diameter size of 0 – 1 m. We'll overwrite our previously-defined variable `predictions`:

```{r}
predictions <- sp_simulate(ref_table = sp_models_info, models = sp_models, 
                           select_sp = c("Albizia saman", "Terminalia mantaly", "Xanthostemon chrysanthus"),
                           extrapolate = c(0,1)) 
head(predictions)
```

## 5. Plot predictions

We can plot out our predictions using `ggplot()`:

```{r message = FALSE, warning = FALSE, fig.width=7, fig.height =2.5, fig.align='center'}
library(ggplot2)

ggplot() +
  facet_wrap(~ species)+ 
  geom_point(data = dplyr::filter(urbantrees, species %in% predictions$species), aes(x = diameter, y = height),
             size=0.35, alpha = 0.9, color = "grey50") +
  
  # prediction interval
  geom_ribbon(data = predictions, aes(x = predictor, ymin = lwr, ymax = upr), 
              alpha = 0.10) +
  
  # regression line
  geom_line(data = predictions, aes(x = predictor, y = fit, lty = extrapolated),
            size = 0.8, alpha = 0.6) +
  
  scale_linetype_manual(values=c("No"= 1 , "Low" = 3, "High" = 3), name = "Extrapolated") +
  
  xlab("Diameter (m)") + ylab("Height (m)") +
  
  coord_cartesian(ylim=c(0, 25), xlim= c(0,1))

```

&nbsp;

Interpretations of the allometric relationships will vary depending on what type of parameter is analysed (e.g. tree height, trunk diameter, age, crown height, crown diameter and leaf area). In our example, we'll have to be cautious in interpretating extrapolated predictions. As you can see in the figure, extrapolations down to a diameter size of zero for *Albizia saman* and *Terminalia mantaly* would not make sense (i.e. the tree cannot have a positive height at that diameter size). Also, while an exponential relationship shows rapid scaling in height for *Terminalia mantaly*, it would have a natural height limit (i.e. would not continue growing taller at such a steep slope). Hence, it is important that allometric relationships are interpreted in conjunction with information on the biology and growth (e.g. environmental and management) conditions associated with the trees.


---

## References

McPherson E. G., van Doorn N. S. & Peper P. J. (2016) Urban Tree Database and Allometric Equations. *General Technical Report PSW-GTR-253, USDA Forest Service*, 86.

Song, X. P., Lai, H. R., Wijedasa, L. S., Tan, P. Y., Edwards, P. J., & Richards, D. R., Height–diameter allometry for the management of city trees in the tropics (in review).

