---
title: "Single-species linear models"
author: 
date: "`r Sys.Date()`"
opengraph:
  image: 
    src: "man/figures/logo.png"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Single-species linear models}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE, warning = FALSE
)
```

```{r include = FALSE, eval = TRUE}
devtools::load_all() # if don't want to use installed version
```

```{r include = FALSE, eval = FALSE}
library(allometree) # if installed (FOR TRAVIS CI)
```

---

## 1. Model selection {#section1}

### For one species

For a particular species of interest, data will be fit to all allometric equations. The best-fit model is selected based on the lowest bias-corrected Aikaike’s information criterion (AICc) value. Here's an example to select the best-fit model for the species *Albizia saman*, using the function `sp_modelselect()`.

```{r sp_modelselect}
data(urbantrees, package = "allometree")

Alb_sam <- urbantrees[urbantrees$species == "Albizia saman", ]  # subset data for one species

results <- sp_modelselect(Alb_sam, 
                          response = "height", predictor = "diameter")  # specify colnames of variables
```

The `results` variable is a list of 3 elements:

1. Table showing models ranked by AICc value, named `all_models_rank`.
2. Best-fit model object, named `best_model`.
3. Table showing information on the best-fit model, named `best_model_info`. More information on the output columns can be found at `?sp_modelselect`.

```{r}
head(results$all_models_rank)
```

```{r}
results$best_model
```

```{r}
results$best_model_info
```


### For multiple species

We can also run `sp_modelselect()` across multiple species in the full dataset `urbantrees`. For this, we use the wrapper function `sp_modelselect_multi()`.

```{r}
results_all <- sp_modelselect_multi(urbantrees, species = "species", # specify colname of species
                                    response = "height", predictor = "diameter")
```

The `results_all` variable is a list of 3 elements:

1. List of tables showing each species' candidate models ranked by AICc value, named `sp_models_rank`.
2. List of each species' best-fit model object, named `sp_models`.
3. Table showing each species' best-fit model information, named `sp_models_info`. 

&nbsp;

An overview of the best-fit models across the `r nrow(results_all$sp_models_info)` species:
```{r}
results_all$sp_models_info
```


## 2. Outlier removal

Let's say we want to check for outlier/influential data points, remove them, and re-fit the filtered dataset to the pre-selected models as defined in `results_all$sp_models_info`. First, lets visualise these outliers, say, for the species *Xanthostemon chrysanthus*.

```{r fig.align='center', fig.width=5.5, fig.height=5.5}
par(mfrow=c(2,2)); plot(results_all$sp_models$`Xanthostemon chrysanthus`)
```
&nbsp;

There are numerous ways to [deal with outliers](http://r-statistics.co/Outlier-Treatment-With-R.html). One way is to remove them based on Cook's Distance (bottom-right plot). In general use, outliers may be defined as points with a Cook's distance more than four times the mean Cook's distance (red line in plot below). Note that this threshold is not fixed, and may be adjusted according to expert knowledge.

```{r, fig.align='center', fig.width=4, fig.height=4}
cooks_dist <- cooks.distance(results_all$sp_models$`Xanthostemon chrysanthus`)
plot(cooks_dist, pch = ".", cex = 2, main = "Outliers by Cook's Distance")
abline(h = 4*mean(cooks_dist, na.rm = T), col = "red")  # threshold line
text(x = 1:length(cooks_dist) + 1, y = cooks_dist, 
     labels = ifelse(cooks_dist > 4 * mean(cooks_dist, na.rm=T), names(cooks_dist), ""), col = "red")  # outlier labels
```

We can create a function `uncooker()` that removes outliers for each species in the dataset `urbantrees`, based on their respective best-fit models in `results_all$sp_models`:

```{r}
uncooker <- function(data, modellist){
  
  result <- data[0,]
  
  for(i in 1:length(modellist)){ # loop over all species 
    
    cooks_dist <- cooks.distance(modellist[[i]]) 
    outliers <- as.numeric(which(cooks_dist > 4*mean(cooks_dist, na.rm=T)))  # row numbers
    subset <- data[(data$species == names(modellist[i])),][-outliers,]
    result <- rbind.data.frame(result, subset)
  }
  return(result)
}

# run the function uncooker()
urbantrees_clean <- uncooker(urbantrees, results_all$sp_models)

nrow(urbantrees_clean)
```

`r nrow(urbantrees) - nrow(urbantrees_clean)` data points were removed to obtain the filtered dataset `urbantrees_clean` (n = `r nrow(urbantrees_clean)`).

```{r include = FALSE}
rm(cooks_dist, uncooker)
```


## 3. Re-fit filtered data

### For one species

We can now fit `urbantrees_clean` to the best-fit equations we defined in [Section 1](#section1). Here's an example for the species *Albizia saman* again, using the function `sp_modelfit()`. This function has an additional argument `modelcode`. In the case of *Albizia saman*, it is `lin_w1` (see `results_all$sp_models_info`). Note that you can also pick any one of the options found in `eqns_info$modelcode`.

&nbsp;

We'll overwrite our previously-defined variables, this time with the filtered dataset:

```{r fit data to pre-selected model, eval = FALSE}
Alb_sam <- urbantrees_clean[urbantrees_clean$species == "Albizia saman", ]

results <- sp_modelfit(Alb_sam, modelcode = "lin_w1", # specify modelcode
                       response = "height", predictor = "diameter") 
```

The new `results` variable is a list of 2 elements:

1. Resulting model object, named `fitted_model`.
2. Table showing information on the resulting model, named `fitted_model_info`.

### For multiple species

Similar to the the wrapper function `sp_modelselect_multi()`, we can also run  `sp_modelfit_multi()` to fit pre-selected models across multiple species. In this function, we need to input a reference table `ref_table` (i.e. `results_all$sp_models_info`) that provides information on the `species` and their corresponding `modelcode`.

```{r}
results_all <- sp_modelfit_multi(urbantrees_clean, 
                                 ref_table = results_all$sp_models_info,
                                 species = "species", # colname in both data & ref_table
                                 modelcode = "modelcode", # colname in ref_table
                                 response = "height", predictor = "diameter")
```

The new `results_all` variable is a list of 2 elements:

1. List of each species' resulting model object, named `sp_models`.
2. Table showing each species' resulting model information, named `sp_models_info`.

&nbsp;

To make our subsequent code less verbose, let's assign the elements `sp_models` and `sp_models_info` to variables with the same name:

```{r}
sp_models <- results_all$sp_models
sp_models_info <- results_all$sp_models_info
```

```{r include = FALSE}
rm(results_all, results, Alb_sam)
```

## 4. Make predictions

We can simulate some data for a species of interest, and use it's model to make predictions. We'll use the species *Albizia saman* again as an example. Let's first simulate the data, based on the range of values for the predictor variable that was used to fit the model:

```{r}
predict_range <- sp_models_info[,c("species","predictor_min", "predictor_max")] # range of predictors
predict_range <- predict_range[predict_range$species == "Albizia saman",]

# simulate 100 data points across range
predict_range_full <- as.data.frame(apply(predict_range, 1, function(x) seq(x["predictor_min"], x["predictor_max"], length.out = 100)))
colnames(predict_range_full) <- predict_range$species
predict_range_full <- tidyr::pivot_longer(predict_range_full, cols = colnames(predict_range_full), names_to = "species", values_to = "predictor")
```

&nbsp;

`sp_predict()` can be used to make predictions on the simulated data `predict_range_full`. This appends the columns `fit` (predicted value), as well as the lower `lwr` and upper `upr` bounds of the prediction interval (default confidence `level` = `0.95`):

```{r}
predictions <- sp_predict(predict_range_full, models = sp_models, ref_table = sp_models_info, predictor = "predictor")
head(predictions)
```

&nbsp;

Note that you can simulate data automatically using the `sp_simulate()` function, which is also a wrapper to `sp_predict()`. This can be done for one or multiple species, using the argument `select_sp`. The simulated data can also be extrapolated beyond the range used to fit the model, using the argument `extrapolate`. Here's an example using the models that we previously derived from the filtered dataset `urbantrees_clean`. Let's say we're interested in both *Albizia saman* and *Hopea odorata*, and overwrite our previously-defined variable `predictions`:

```{r}
predictions <- sp_simulate(ref_table = sp_models_info, models = sp_models, 
                           selected_spp = c("Albizia saman", "Hopea odorata"),
                           extrapolate = c(0,3)) # extrapolate predictor variable down to 0 and up to 3m
head(predictions)
```

## 5. Plot predictions



---

## References

McPherson E. G., van Doorn N. S. & Peper P. J. (2016) Urban Tree Database and Allometric Equations. *General Technical Report PSW-GTR-253, USDA Forest Service*, 86.

Song, X. P., Lai, H. R., Wijedasa, L. S., Yee, A. T. K., Tan, P. Y., Richards, D. R., Streamlining management practices based on the size allometry of tropical street trees (in prep).

